---
title: "GO Analysis"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Purpose of this code
To annotate datasets with GO Terms and then give all Terms present a p-value based on Fisher's Exact Test. There are four data sets and three GO categories, furthermore the datasets have been analysed for all genes, up-regulated only and down-regulated only to determine components/functions/processes which are being promoted and those which are being reduced in expression. Due to the number of processes necessary to process the data like this, the code takes quite a long period of time to run, approx. 1-2 hours. An alternative final script for testing is provided as explained below at appropriate step.

After code is run, open one of the csv's and copy all text except the top row. Paste into the text box at http://revigo.irb.hr/ and click Start Revigo. Select Treemap from the tabs.

It is expected the datasets will be in your working directory having previous run the last .Rmd file as explained in comments.

## Packages

```{r echo = FALSE}
library(readr) #loading tsvs
library(tibble) #using tibbles for df's
library(dplyr) #manipulating data
```

## Load Datasets
Adding relevant files from working directory to working session. Please set working directory if necessary to where all files are.
gene2go MUST be unzipped first in folder.
```{r echo = FALSE}
setwd() <- getwd() #Input location of files on computer
Old_baseline_2h_results <- read.csv("Old_baseline_2h_results.csv",header = T)
Old_baseline_trained_results <- read.csv("Old_baseline_training_results.csv",header = T)
YG_baseline_2h_results <- read.csv("YG_baseline_2h_results.csv",header = T)
YG_baseline_trained_results <- read.csv("YG_baseline_training_results.csv",header = T)
gene2go <- read_delim("~/Downloads/gene2go (1)", 
          "\t", escape_double = FALSE, trim_ws = TRUE)
```

## Function
This will be used in the next step to create individual p-values for GO Terms present in a file put in the working directory. Explanation of output file naming convention is in the comments.
```{r pressure, echo=FALSE}
#Instructions for general use
#Please ensure the files to analyze are in the working directory.
#This is designed to be run after "Group Project Code.Rmd" but can be used for other purposes if tailored correctly
#The expected format is columns: GeneID (numerical), Baseline_mean, Trained_mean, theta, prob, log2FC, threshold
#HOWEVER, only GeneID, log2FC and prob (probability) are actually used so others are superfluous
#In some cases GeneID may have the column name X1, in this case it is changed to GeneID in the function
#GeneID should be column 1

#Also you must have, in the working directors, a file called gene2go which indexes all the GO Terms for each gene. 
#It is available from ftp://ftp.ncbi.nih.gov/gene/DATA/gene2go.gz OR provided on our github - IT MUST BE UNZIPPED

#Function for analyzing GO terms based on their category
#Options - all must be completed and in quotes except the Dataset
#Dataset = insert name of dataset to analyse, this needs to be loaded in the Working Session before running.
#Age = "O" for Old OR "Y" for Young (with quotes)
#Comparison = "bv2" for baseline vs 2 hours training OR "bvt" for baseline vs training for 2 weeks (with quotes)
#Category = "C" for Component (location in cell) OR "F" Function (function of protein from gene) OR "P" for Process (biological process it is involved with) (with quotes)
#Regulated = "Up" for upregulated DEG's only, "Down" for down regulated and "UpAndDown" for both

#Output - a list of all unique GO Terms which are significantly represented in the input file
## FIle name is systematically named with the following components:
## C/F/P (GO Category) + O/Y (age group) + M- (Male) + bv2/bv2 (baseline vs 2 hours/trained) + U&D/U/D Up & Down/Up/Down Regulated

#First Source the GO Analysis Function

GO.Analyzer <- function(Dataset, Age = c("O", "Y"), Comparison = c("bv2", "bvt", "2vt"), GO.Category = c("C", "F", "P"), Regulated = c("Up", "Down", "Both")){
  
  #Obtain dataset from function
  data <- Dataset
  
  if(Regulated == "Up"){
    data <- data[which(data$log2FC > 0),]
  }
  if(Regulated == "Down"){
    data <- data[which(data$log2FC < 0),]
  }
  if(Regulated == "Both"){
    Regulated <- "UpAndDown"
  }
  
  #Rename first column if necessary
  if(colnames(data[1]) == "X1"){
    data <- dplyr::rename(data, GeneID = X1)
  }
  if(colnames(data[1]) == "X"){
    data <- dplyr::rename(data, GeneID = X)
  }
  #Check the column is now named correctly, if not error message
  if(colnames(data[1]) != "GeneID"){
    print("GeneID may not be named correctly in tibble data")
  }
  
  #Define GO.Category
  if(GO.Category == "C"){
    GO.Category <- "Component"
  }
  if(GO.Category == "F"){
    GO.Category <- "Function"
  }
  if(GO.Category == "P"){
    GO.Category <- "Process"
  }

  #filter gene2go for just homo sapiens (tax id = 9606)  then add gene2g0 for annotations
  gene2go <- gene2go[which(gene2go$`#tax_id` == 9606),]
  annotations <- gene2go
  
  # combine data files for differentially expressed genes and gene annotation
  data.annotated <- merge(x = data, y = annotations, by = "GeneID")
  
  # Make a list of all pathways/GO Terms etc - filter first
  #Filter by Category of GO Terms - Component, Function or Process
  filtered_results <- annotations[which(annotations$Category == GO.Category),]
  filtered_results <- as.data.frame(filtered_results)
  
  #Create a list of unique GO terms in order to go through in loop
  GO_BP<-filtered_results$GO_ID
  uniGO_BP <- unique(GO_BP)
  uniGO_BP <- sort(uniGO_BP)
  
  #Create a list of all GO Terms for genes which are significantly expressed
  Sig.Genes<-subset(data.annotated,data.annotated$prob>0.8)

  # Run a Fisher test on all GO Terms, one by one in a loop
  #First create a table to present results of Fisher test
  summaryresults<-data.frame(Functionname=as.character(),pvalue=as.numeric(),sigInID=as.numeric(),nonSigInID=as.numeric(),sigNotInID=as.numeric(),nonSigNotInID=as.numeric())
  #Loop through all GO Terms in data set
  #Then define the Go Term to be looked at
  #Create Fisher Matrix
  #In first cell number of Genes with our GO Term and also significant
  #Down a cell the number that have the GO Term but are not significant
  #Top right cell is those that are significant but don't have the GO Term
  #Down a cell is the number of all the Genes without the GO Term and that are not significant
  #Run Fisher Test and combine results to create summary results
  for (i in 1:length(uniGO_BP)){
    GO.Term <- data.annotated[data.annotated$GO_ID == uniGO_BP[i], ] 
    FisherMatrix <- matrix(nrow=2, ncol=2, data=c(-1, -2, -3, -4), dimnames=list("Genes including GO Term" = c("y", "n"), "Genes with sign pvalue" = c("y", "n")))
    FisherMatrix[1,1]  =  length(intersect( GO.Term$GeneID,  Sig.Genes$GeneID) ) 
    FisherMatrix[2,1]   = length(setdiff(Sig.Genes$GeneID, GO.Term$GeneID ))
    FisherMatrix[1,2]   = length(setdiff(GO.Term$GeneID, Sig.Genes$GeneID  ))
    FisherMatrix[2,2]   = length(setdiff(data.annotated$GeneID, union(GO.Term$GeneID,Sig.Genes$GeneID)))
    results<-fisher.test(FisherMatrix, alternative = "greater")
    output<-data.frame(GO.Name=uniGO_BP[i],pvalue=results$p.value,sigInID=FisherMatrix[1,1],nonSigInID=FisherMatrix[1,2],sigNotInID= FisherMatrix[2,1],nonSigNotInID=FisherMatrix[2,2])
    summaryresults<-rbind(summaryresults,output)
    if (i == 1){
    }
  }
  
  #Add an addjust p value by false discovery rate (fdr)
  summaryresults$p.adjusted<-p.adjust(summaryresults[,2], method="fdr")
  sig.list<-subset(summaryresults,summaryresults$p.adjusted<0.05)
  
  #Create output file with GO Terms and their adjusted p-value
  dplyr::select(summaryresults, GO.Name, p.adjusted) %>% filter( p.adjusted < 0.05) %>% write_delim(paste(GO.Category, Age, "M-", Comparison, Regulated, "Regulated.tsv", sep = ""))
}
```

## Loop Function
The section where the function is run. This takes some time to complete, the next chunk of code is for testing purposes.
```{r}
#LOOP FOR CREATING GO TSVs
#Load datasets under simple, standardized names
D1 <- YG_baseline_2h_results
D2 <- YG_baseline_training_results
D3 <- Old_baseline_2h_results
D4 <- Old_baseline_training_results

#Loop for Up&Down, Up and Down regulation. This gives three separate outputs for each, allows further analysis on how the GO Terms are expressed.
#Comment provided only for first if statement as they are the same just for each dataset as defined above
for (r in 1:3){
  regulist <- c("UpAndDown", "Up", "Down")
 
  #Loops through each dataset 
  for (d in 1:4){
  
    if (d == 1){
      Dataset <- D1
      Age <- "Y" #Defines the age categoery
      Comparison <- "bv2" #defines the comparison, in this case baseline vs 2 hours
      GO.Category = c("C", "F", "P") #defines the GO Term category Component/Function/Process
      
      #Loops through each GO Term category
      for (c in 1:3){
        GO.Analyzer(Dataset, Age, Comparison, GO.Category[c], regulist[r]) #defines all terms to put into the function
      }
    }
      if (d == 2){
        Dataset <- D2
        Age <- "Y"
        Comparison <- "bvt"
        GO.Category = c("C", "F", "P")
        
        for (c in 1:3){
          GO.Analyzer(Dataset, Age, Comparison, GO.Category[c], regulist[r])
        }
      }   
      if (d == 3){
        Dataset <- D3
        Age <- "O"
        Comparison <- "bv2"
        GO.Category = c("C", "F", "P")
        
        for (c in 1:3){
          GO.Analyzer(Dataset, Age, Comparison, GO.Category[c], regulist[r])
        }
      }
      if (d == 4){
        Dataset <- D4
        Age <- "O"
        Comparison <- "bvt"
        GO.Category = c("C", "F", "P")
          
          for (c in 1:3){
            GO.Analyzer(Dataset, Age, Comparison, GO.Category[c], regulist[r])
          }
        }
      }
    } 
```

## Loop Function - shortened version for testing
This is the same as above but shortened for testing purposes
```{r}
#LOOP FOR CREATING GO TSVs
#Load datasets under simple, standardized names
D1 <- YG_baseline_2h_results

#Loop for Up&Down, Up and Down regulation. This gives three separate outputs for each, allows further analysis on how the GO Terms are expressed.
#Comment provided only for first if statement as they are the same just for each dataset as defined above
for (r in 1:3){
  regulist <- c("UpAndDown", "Up", "Down")
 
  #Loops through each dataset 
  for (d in 1:4){
  
    if (d == 1){
      Dataset <- D1
      Age <- "Y" #Defines the age categoery
      Comparison <- "bv2" #defines the comparison, in this case baseline vs 2 hours
      GO.Category = c("C", "F", "P") #defines the GO Term category Component/Function/Process
      
      #Loops through each GO Term category
      for (c in 1:1){
        GO.Analyzer(Dataset, Age, Comparison, GO.Category[c], regulist[r]) #defines all terms to put into the function (TESTING - only looks at Component)
      }
    }
  }
} 
```
